Linear Reduction: Parsing and Interpreting Code Without Trees

### 2024-12-04 Paper/paper.tex
Updated paper
Added bibliography

%\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, mathtools}
\usepackage{enumitem}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{graphicx}

@@

\begin{abstract}
test
    In this paper we discuss an alternative method to parsing and interpreting human-readable code.
    This method utilizes an iterative approach, as opposed to a recursive tree-oriented approach found in most modern compilers and interpreters.
    We implement our algorithm in an interpreter and compare it with an interpreter created using more classical means.

@@

Nowadays, most programming language interpreters follow the following pipeline:
\begin{enumerate}
    \item Lexing: lexically analyze the input code, separating it into atomic \textit{tokens}.
    \item Parsing: convert the parsed tokens into a \textit{parse tree}: a tree which represents the grammar of the code.
    \item Interpreting: recursively descend the tree and execute the code according to the instructions embedded in the tree.
\end{enumerate}
For example, we can parse the following expression:

@@

Unfortunately, constructing such a tree is not so easy.
Multiple methods for doing so exist, as well as tools for automating this process.

We propose a novel method for the process of interpretation, which doesn't use parse trees.
Parsing code generated by context-free grammars (CFGs) is inherently a recursive process, and thus it is not surprising that our algorithm can be altered to instead produce parse trees.

@@

We implement our algorithm in an interpreter for a toy language we dub ``LLang''.
We also create an interpreter for LLang using menhir\NEEDCITE{}.
In this section we detail the implementation of our algorithm and compare and contrast the two implementations.

@@

\noindent{\bf If Statements}
\blist
    \item $\iif\ \tsig \longto \ifbool\ \fst\ (\_,n,s\to n,\epsilon,s+[\{\mapsto\Lbrace\a,(\mapsto\lparen\a)]$
    \item $\ifbool\ \code \longto \ifthen\ \fst\ (n,\xi,s\to(n,\xi),\epsilon,s)$
    \item $\ifthen\ \code \longto \epsilon\ \_\ ((n,\xi_1),\xi_2,s\to\varnothing,(n=0?\ \xi_2:\xi_1),\pop s)$
\elist

@@

In this section we compare the runtimes of both implementations (one using linear reduction and one using menhir\NEEDCITE{}).
A few notes beforehand:
\begin{itemize}
    \item The linear reduction implementation has more features than the menhir implementation, this should not severely impact the comparisons and we will of course not be comparing the implementations
            on code which the menhir implementation does not support.
    \item LLang supports left composition.
            Meaning code of the form \ttt{f x y} is read as \ttt{(f x) y}.
            We found no way of supporting this with menhir.
            This is because our grammar for the menhir implementation contains the following fragment:
            \[ {\it value} \ccoloneqq {\it const}\gbar{\it ident}\gbar{\it funcall},\qquad {\it funcall}\ccoloneqq{\it value}\ {\it value} \]
            So \ttt{f x y} can be parsed as \ttt{f (x y)} instead:

@@

>>>>

%\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, mathtools}
\usepackage{enumitem}
\usepackage{float}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{graphicx}
\usepackage[backend=biber]{biblatex}

\addbibresource{paper-bib.bib}

@@

\begin{abstract}
    In this paper we discuss an alternative method to parsing and interpreting human-readable code.
    This method utilizes an iterative approach, as opposed to a recursive tree-oriented approach found in most modern compilers and interpreters.
    We implement our algorithm in an interpreter and compare it with an interpreter created using more classical means.

@@

Nowadays, most programming language interpreters follow the following pipeline (see \cite{dragon})
\begin{enumerate}
    \item Lexing: lexically analyze the input code, separating it into atomic \textit{tokens}.
    \item Parsing: convert the parsed tokens into a \textit{parse tree}: a tree which represents the grammar of the code.
    \item Interpreting: recursively descend the tree and execute the code according to the instructions embedded in the tree.
\end{enumerate}
For example, we can parse the following expression:

@@

Unfortunately, constructing such a tree is not so easy.
Multiple methods for doing so exist (some are listed in \cite{dragon}), as well as tools for automating this process.

We propose a novel method for the process of interpretation, which doesn't use parse trees.
But our algorithm is not limited to this: parsing code generated by context-free grammars (CFGs) is inherently a recursive process, and thus it is not surprising that our algorithm can be altered to
instead produce parse trees.

@@

We implement our algorithm in an interpreter for a toy language we dub ``LLang''.
We also create an interpreter for LLang using menhir\cite{menhir}.
In this section we detail the implementation of our algorithm and compare and contrast the two implementations.

@@

\noindent{\bf If Statements}
\blist
    \item $\iif\ \tsig \longto \ifbooll\ \fst\ (\_,n,s\to n,\epsilon,s+[\{\mapsto\Lbrace\a,(\mapsto\lparen\a)]$
    \item $\ifbooll\ \code \longto \ifthen\ \fst\ (n,\xi,s\to(n,\xi),\epsilon,s)$
    \item $\ifthen\ \code \longto \epsilon\ \_\ ((n,\xi_1),\xi_2,s\to\varnothing,(n=0?\ \xi_2:\xi_1),\pop s)$
\elist

@@

In this section we compare the runtimes of both implementations (one using linear reduction and one using menhir\cite{menhir}).
A few notes beforehand:
\begin{itemize}
    \item The linear reduction implementation has more features than the menhir implementation, this should not severely impact the comparisons and we will of course not be comparing the implementations
            on code which the menhir implementation does not support.
    \item LLang supports left composition.
            Meaning code of the form \ttt{f x y} is read as \ttt{(f x) y}.
            We found no way of supporting this with menhir.
            This is because our grammar for the menhir implementation contains the following fragment:
            \[ {\it value} \ccoloneqq {\it const}\gbar{\it ident}\gbar{\it funcall},\qquad {\it funcall}\ccoloneqq{\it value}\ {\it value} \]
            So \ttt{f x y} can be parsed as \ttt{f (x y)} instead:

@@

\printbibliography
